---
title: "ass1"
author: "Arho Toikka"
date: "17 3 2021"
output: html_document
---

```{r}
# 2. basic computer skills
library(tidyverse)

m <- 0.2
sd <- 0.01
alpha <- m * (m*(1-m)/sd-1)
beta <- alpha*(1-m)/m

ggplot(data = data.frame(x = c(0, 1)), aes(x)) +
  stat_function(fun = dbeta, n = 101, args = list(shape1=alpha, shape2=beta))

d <- data.frame(x=rbeta(1000, alpha,beta))

ggplot(data = data.frame(x = c(0, 1)), aes(x)) +
  geom_histogram(data=d, aes(x=x, y=..density..), binwidth = 0.02, alpha=0.2) +
  stat_function(fun = dbeta, n = 101, args = list(shape1=alpha, shape2=beta))

mean(d$x)
var(d$x)
quantile(d$x, 0.025)
quantile(d$x, 0.975)


```
# 3. Bayes
A group of researchers has designed a new inexpensive and painless test for detecting lung cancer. The test is intended to be an initial screening
test for the population in general. A positive result (presence of lung cancer) from
the test would be followed up immediately with medication, surgery or more extensive and expensive test. The researchers know from their studies the following
facts:

Test gives a positive result in 98% of the time when the test subject has lung
cancer.
 Test gives a negative result in 96 % of the time when the test subject does
not have lung cancer.
 In general population approximately one person in 1000 has lung cancer.
The researchers are happy with these preliminary results (about 97% success rate),
and wish to get the test to market as soon as possible. How would you advise them?
Base your answer on Bayes’ rule computations.

P(posi) = P(posi|cancer=0.98)*P(cancer=0.001)+P(posi|no cancer=0.04)*P(no cancer=0.999) = 0.04094

P(cancer|posi) = P(posi|cancer=0.98)*P(cancer=0.001) / 0.04094 = 0.024 - too many false positives given low prevalence.

P(cancer and positive) = (0.98*0.001)/((0.98*0.001)+(0.02*0.001)+(0.96*0.999)+(0.04*0.999)) = 0.098%

# 3. bayes functions

We have three boxes, A, B, and C. There are
 2 red balls and 5 white balls in the box A,
 4 red balls and 1 white ball in the box B, and
 1 red ball and 3 white balls in the box C.


```{r}


#test 
practice_boxes <- matrix(c(2,2,1,5,5,1), ncol = 2,
dimnames = list(c("A", "B", "C"), c("red", "white")))

# actual
real_boxes <- matrix(c(2,4,1,5,1,3), ncol = 2,
dimnames = list(c("A", "B", "C"), c("red", "white")))


```
Consider a random experiment in which one of the boxes is randomly selected and
from that box, one ball is randomly picked up. After observing the color of the
ball it is replaced in the box it came from. 

Suppose also that on average box A is selected 40% of the time and box B 10% of the time (i.e. P(A) = 0.4).

a) What is the probability of picking a red ball?


```{r}
probabilities <- c(A=0.4, B=0.1, C=0.5)

p_red <- function(boxes, probabilities) {
  p_red <- (boxes["A", "red"]/sum(boxes["A",]))*probabilities["A"]+
    (boxes["B", "red"]/sum(boxes["B",]))*probabilities["B"]+
    (boxes["C", "red"]/sum(boxes["C",]))*probabilities["C"]
  return(unname(p_red))
}

p_red(practice_boxes, probabilities)
```

b) If a red ball was picked, from which box it most probably came from?

P(cancer|posi) = P(posi|cancer=0.98)*P(cancer=0.001) / 0.04094 = 0.024
P(A|red) = P(red|)*P(A)/P(Red)
```{r}
p_box <- function(boxes, probabilities) {
  p_red_A <- ((boxes["A", "red"]/sum(boxes["A",]))*probabilities["A"]) 
  p_red_B <- ((boxes["B", "red"]/sum(boxes["B",]))*probabilities["B"])
  p_red_C <- ((boxes["C", "red"]/sum(boxes["C",]))*probabilities["C"])
  p_a_red <- p_red_A / (p_red_A+p_red_B+p_red_C)
  p_b_red <- p_red_B / (p_red_A+p_red_B+p_red_C)
  p_c_red <- p_red_C / (p_red_A+p_red_B+p_red_C)
  unname(c(p_a_red, p_b_red, p_c_red))
}
p_box(practice_boxes, probabilities)


```
# 5. 

Assume that on average fraternal twins (two fertilized eggs and
then could be of different sex) occur once in 150 births and identical twins (single
egg divides into two separate embryos, so both have the same sex) once in 400
births (Note! This is not the true values, see Exercise 1.6, page 28, in BDA3).
American male singer-actor Elvis Presley (1935 – 1977) had a twin brother who
died in birth. 
Assume that an equal number of boys and girls are born on average.
What is the probability that Elvis was an identical twin? 
Show the steps how you
derived the equations to compute that probability.

P(fr_twin = 1/150)
P(id_twin = 1/400)
P(boy|id_twin=1)
P(boy|fr_twin=0.5)
P(id_twin|boy) = P(boy|id_twin)*P(id_twin) / (P(boy|id_twin)*P(id_twin) + P(boy|fr_twin)*P(fr_twin))

Implement this as a function in R that computes the probability


```{r}
p_identical_twin <- function(fraternal_prob=1/125, identical_prob=1/300) {
  identical_prob / (identical_prob + 0.5*fraternal_prob)
}
```   

