---
title: "BDA - Assignment 5"
author: "Anonymous"
output: html_document 
---


```{r, include=FALSE}
# This chunk sets echo = TRUE as default, that is print all code.
# knitr::opts_chunk$set can be used to set other notebook generation options, too.
# include=FALSE inside curly brackets makes this block not be included in the pdf.
knitr::opts_chunk$set(echo = TRUE)
```

# Loaded packages


```{r}
# To install aaltobda, see the General information in the assignment.
library(aaltobda)

library(markmyassignment)
assignment_path <-
  paste("https://github.com/avehtari/BDA_course_Aalto/,
blob/master/assignments/tests/assignment5.yml", sep="")
set_assignment(assignment_path)

```



# Exercise 1

We implement the Metropolis algorithm as an R function for the bioassay data using the
Gaussian prior as in Assignment 4, that is

$$
(\alpha, \beta) \sim N(
\begin{pmatrix}0 \\ 10\end{pmatrix}, 
\begin{pmatrix}2^2 & 12\\12 & 10^2\end{pmatrix})
$$


```{r}
mu <- c(0, 10)
S <- matrix(c(2^2, 2*10*0.6, 
              2*10*0.6, 10^2 ),
            ncol = 2)
```

## a)

We start by implementing a function called density_ratio to compute the density ratio function (*r* in Eq. (11.1) in BDA3).

The unnormalised posterior is an exponent of the sum of log likelihood and log prior

```{r}

unnormalised_posterior <- function(a, b, x, y, n) {
  loglik <-   bioassaylp(a = a, 
                         b = b,
                         x = x, 
                         y = y, 
                         n = n)
  
  prior <- dmvnorm(c(a, b), mu, S, log = TRUE)
  
  exp(loglik + prior)
  
}
```

The density ratio is the ratio of the unnormalised posterior at the previous and proposed values of $\alpha$ and $\beta$.

```{r}

density_ratio <- function(alpha_propose,
                          alpha_previous,
                          beta_propose, 
                          beta_previous,
                          x,
                          y, 
                          n) {
  
  prop <-   unnormalised_posterior(a = alpha_propose, 
                                   b = beta_propose,
                                   x = x, 
                                   y = y, 
                                   n = n)
  
  prev <- unnormalised_posterior(a = alpha_previous, 
                                 b = beta_previous,
                                 x = x, 
                                 y = y, 
                                 n = n)
  
  prop / prev
}
```


## b)

We implement a function called Metropolis_bioassay() which implements the Metropolis algorithm using the density_ratio(). We utilise simple (normal) proposal distributions. 

```{r}
Metropolis_bioassay <- function(bioassay,
                                init_a, 
                                init_b, 
                                iters = 4000,
                                a_prop_scale = 1,
                                b_prop_scale = 5) {
  
  res <- data.frame(alpha = rep(as.numeric(NA), iters),
                    beta = rep(as.numeric(NA), iters),
                    accept = NA)
  
  a_prev <- init_a
  b_prev <- init_b
  
  for (i in 1:iters) {
    
    a_prop <- rnorm(1, a_prev, a_prop_scale)
    b_prop <- rnorm(1, b_prev, b_prop_scale)
    
    r <- density_ratio(alpha_propose = a_prop,
                       alpha_previous = a_prev,
                       beta_propose = b_prop,
                       beta_previous = b_prev,
                       x = bioassay$x, 
                       y = bioassay$y, 
                       n = bioassay$n)
    accept <- runif(1) < r
    res[i, "alpha"] <- ifelse(accept, a_prop, a_prev)
    res[i, "beta"] <- ifelse(accept, b_prop, b_prev)
    res[i, "accept"] <- accept
    
    a_prev <- res[i, "alpha"]
    b_prev <- res[i, "beta"]
    
  }
  
  res
}

```

We run the Metropolis algorithm using the scales 1 and 5 for the alpha and beta proposal distributions, respectively
```{r}
bioassay <- get(data("bioassay"))
df <- Metropolis_bioassay(bioassay,
                          init_a = rnorm(1, 1, 3),
                          init_b = rnorm(1, 10, 5))
                          
```

The acceptance rate is `r mean(df$accept)` which is close to the optimal 40%. Therefore we make no adjustments.


# Exercise 2

## a) 

*Describe the basic idea of the Metropolis algorithm*.

The metropolis algorithm can be seen as a stochastic mode finding algorithm which converges to the target distribution. 

## b)

