---
title: "BDA - Assignment 2"
author: "Anonymous"
output: 
  html_document: 
  toc: yes
toc_depth: 1
---


```{r setup, include=FALSE}
# This chunk sets echo = TRUE as default, that is print all code.
# knitr::opts_chunk$set can be used to set other notebook generation options, too.
# include=FALSE inside curly brackets makes this block not be included in the pdf.
knitr::opts_chunk$set(echo = TRUE, comment = NA)
```

```{r}
# To install aaltobda, see the General information in the assignment.
library(aaltobda)

```

# Exercise 2

```{r} 
algae <- get(data("algae")) 
n <- length(algae)
y <- sum(algae)
a <- 2
b <- 10
```

The model used in this excercise is

$$
p(\pi) \sim Beta(2, 10) \\
p(y \mid \pi) \sim Binom(n, y, \pi)
$$

and the data are

```{r}
table(algae)
```


## a)

It is easy to see that the following holds:

$$
p(\pi) \propto \pi^{\alpha} ( 1 - \pi)^{\beta}, \\
p(y \mid \pi) \propto \pi^{y} ( 1 - \pi)^{(n - y)}, \\
p( \pi \mid y) \propto \pi^{y + \alpha} ( 1 - \pi)^{(n - y + \beta)},
$$

where the posterior has the form of the Beta distribution with parameters $\alpha + y$ and $n -y + \beta$.

In this case 

- $n$ = `r n`
- $y$ = = `r y`
- $\alpha$ = `r a`
- $\beta$ = `r b`

And so 

$$
p( \pi \mid y) \sim Beta(46, 284)
$$

## b)

The expectation of the beta distribution with parameters $\alpha$ and $\beta$ is $\alpha / (\alpha + \beta)$.

```{r}

# parameters of beta posterior with beta prior a, b and bibomial likelihood 
# with data x
betabin_posterior <- function(a, b, x) {
  n <- length(x)
  y <- sum(x)
  c("a" = a + y, "b" = b + n - y)
}
beta_point_est <- function(prior_alpha, prior_beta, data) {
  th <- betabin_posterior(prior_alpha, prior_beta, data)
  th["a"] / (th["a"] + th["b"])
}

```

In the case of $p( \pi \mid y)$, the expectation corresponds to

```{r}
beta_point_est(a, b, algae)
```

A 90% posterior interval can be obtained from the quantiles of the posterior Beta distribution:

```{r}

beta_interval <- function(prior_alpha, prior_beta, data, prob  = 0.9) {
  th <- betabin_posterior(prior_alpha, prior_beta, data)
  q1 <- (1 - prob) / 2
  q2 <- prob + q1
  qbeta(c(q1, q2), th["a"], th["b"])
}

beta_interval(a, b, algae, 0.9)
```

## c)

We can use the beta cumulative distribution function to evaluate the probability $P(\pi < 0.2 \mid y) = F_{\pi \mid y}(0.2)$.

In this case, the probability is

```{r}
beta_low <- function(prior_alpha, prior_beta, data, pi_0 = 0.2) {
  th <- betabin_posterior(prior_alpha, prior_beta, data)
  pbeta(pi_0, th["a"], th["b"])
}
beta_low(2, 10, algae, 0.2)
```

## d)

We assume the measurements $y_i$ are independent and idetically distributed.

## e)

```{r}
plot_beta_posterior <- function(a, b, data) {
  th <- betabin_posterior(a,b,data)
  x <- seq(0, 1, by = 0.001)
  y <- dbeta(x, th["a"], th ["b"])
  plot(x, y, xlim = c(0, 1), ylim = c(0, 20), type = "l",
       xlab = "pi", ylab = "Density")
  title(paste0("a =", a, " b=", b, " y=", sum(data), " n=", length(data)))
}
```

There is quite a lot of data so reasonably uninformative or vague priors do not have a significant effect on the posterior. 
A very informative prior will however affect inference.


```{r}
par(mfrow = c(2,2))
plot_beta_posterior(0.001,0.001, algae)
plot_beta_posterior(1,5, algae)
plot_beta_posterior(2,10, algae)
plot_beta_posterior(50,50, algae)
```


